---
title: "Project 1"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Introduction

In this project, you’re given a text file with chess tournament results where the information has some structure. Your
job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database)
with the following information for all of the players:

    Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents  
    
For the first player, the information would be:
    
    Gary Hua, ON, 6.0, 1794, 1605  

1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and
dividing by the total number of games played.  

If you have questions about the meaning of the data or the results, please post them on the discussion forum. Data
science, like chess, is a game of back and forth...  

The chess rating system (invented by a Minnesota statistician named Arpad Elo) has been used in many other contexts,
including assessing relative strength of employment candidates by human resource departments.

# Solution

## Read tournament data

Load input data from Github. 

```{r load-data}
input_data <- read.csv("https://raw.githubusercontent.com/himalayahall/DATA607/main/Project1/tournamentinfo.txt")
head(input_data)
```


## Data preparation

Unlist for easier downstream processing.

```{r}
input_data <- unlist(input_data, use.names = FALSE)
head(input_data)
```

Let's convert all data to uppercase to ensure mixed casing in the input does not break downstream processing.

```{r}
input_data <- unlist(
                    map(input_data, 
                        function(x) 
                            str_to_upper(x)), 
                    use.names = FALSE)
```

Observe that first 3 rows are headers and don't contain player info.

```{r}
head(input_data, n = 10)
```

Based on this observation, we skip the 3 header rows.

```{r skip-header-rows}
input_data <- input_data[-1:-3]
head(input_data, n = 6)
```

Our next observation is that data for each player is provided on 2 consecutive rows, followed by a dashed separator line. 1^st^ row contains the player name, total points, and games played. The 2^nd^ row contains the State and initial rank.  

Based on this observation, let's split the input matrix into 2 components.  

The 1^st^ component will contain player name, total points, and games played. We can extract this data by scooping up every 3^rd^ row from the input matrix, starting at row 1 .  

```{r extract-players-and-games}
mPlayersAndGames <- unlist(input_data[seq(1, length(input_data), 3)], use.names = FALSE)
```

A quick look shows we have extracted data correctly.

```{r}
mPlayersAndGames[1:5]
```

The 2^nd^ component will contain player State and pre-rating. This time we start at row 2 and scooping up every 3^rd^ row from the input matrix.

```{r extract-states-and-ranks}
mStatesAndRanks <- unlist(input_data[seq(2, length(input_data), 3)], use.names = FALSE)
```

Once again, a quick look shows data has been extracted correctly.

```{r}
mStatesAndRanks[1:5]
```

## Generate static player data

At this point we have the necessary components to generate static data for each player - i.e. id, name, state, total points, and pre-rating.  

```{r}
player_id <- as.integer(
                str_extract(mPlayersAndGames, '\\d+'))

player_name <- str_trim(                        # assume player names start with a letter
                str_extract(mPlayersAndGames, "[A-Z][^\\|]+")) 

player_state <- str_extract(                    # assume 2-letter abbreviation for State
                mStatesAndRanks, "[A-Z][A-Z]")  

player_total_points <- as.numeric(str_extract(mPlayersAndGames, "[0-9]+\\.[0-9]"))

player_prerating <- as.numeric(str_remove(str_extract(mStatesAndRanks, "R:[ ]+[0-9]{1,}"), "R:[ ]+"))

```

It's a good idea to spot check our static data by comparing with the original.  

Let's print some values from both the original and extracted static data to compare visually.  

    Notice that all player names have been successfully extracted. For example, a complex name like SOFIA ADINA STANESCU-BELLU has been extracted.    
    
    Also notice that pre-rank for ZACHARY has been successfully extracted although it was 'corrupted' (1220P13) in original input.  

```{r}
test_player_idx <- c(1, 12, 15, 22, 28, 58, 64)
sanity_check <- data.frame(
                    player_id[test_player_idx],
                    player_name[test_player_idx],
                    player_total_points[test_player_idx],
                    player_prerating[test_player_idx]
)
colnames(sanity_check) <- c('id', 'name', 'total_points', 'pre_rank')

sanity_check
```

In the extracted data ZACHARY JAMES HOUGHTON has 4.5 total points and 1220 pre-rank, and SOFIA ADINA STANESCU-BELLU has 3.5 total points and 1507 pre-rank.  

Let's compare the extracted name and total points for ZACHARY JAMES HOUGHTON and SOFIA ADINA STANESCU-BELLU against the original. It checks out!

```{r}
sz_idx <- c(15, 28) 
mPlayersAndGames[sz_idx]
```

Finally, we verify that extracted State and pre-rank data matches original input. Everything checks out.

```{r}
mStatesAndRanks[sz_idx]
```

## Missing opponent id

The following codes may appear in a chess cross table:

    W - win, worth 1 point  
    
    L - lose, worth 0 points  
    
    D - draw, worth 0.5 points  
    
    B - full point bye, worth 1 point (given to the left-over player when there are an odd number of players in a tournament round)  
    
    H - half point bye, worth 0.5 points (players can request these when they know they won't be able to make it to certain rounds in a tournament. They are normally only available in the first few rounds of a tournament, and tournament directors often limit a player to a small number of bye requests)  
    
    X - win by forfeit, worth 1 point  
    
    F - lose by forfeit, worth 0 points (and usually results in automatic withdrawal from the rest of the tournament)  
    
    U - unplayed game, worth 0 points (in a round robin, this shows up for any games that haven't been played yet; in a Swiss tournament, this would show up for games following a forfeit loss. This could also show up in a situation where a player requests more byes than the tournament director permits—the director could allow the player to miss the games without withdrawing from the tournament, but the player would score no points for the missed games)  

We observe that byes (B, H), forfeits (F, X), and unplayed games (U) will not contain opponents ids.  

For example, there is no id of the opposing player for game 5 played by KENNETH J TACK, game 6 played by VIRAJ MOHILE does not contain the opposing player id and games 2-7 do not contain opposing player id for ASHWIN BALAJI.

```{r}
kva_idx <- c(12, 58, 62)    # index for KENNETH, VIRAJ, ASHWIN
mPlayersAndGames[kva_idx]

```

To make downstream processing more robust, let's replace all missing opposing player ids with NA. 

```{r replace-missing-ids}
mPlayersAndGames <- mPlayersAndGames %>% 
                        str_replace_all("\\|([BHFXU])([ \t\f\n])+", "|\\1\\2 \\N\\A")
```

Validate that missing player ids have been replaced with NA.

```{r}
mPlayersAndGames[kva_idx]

```

## Data preparation for computing average opponent ranking

Extract opposing player ids into a flattened list of integers. 

```{r warning = FALSE}
p_opponent_ids <- mPlayersAndGames %>% 
                    str_extract_all("[A-Z][ ]+([0-9]+|NA)") %>%  # extract game result 
                                                                 # and opponent player id
                    unlist %>%                          # flatten
                    str_remove("[A-Z][ ]+") %>%         # extract player id (remove game result)
                    as.integer()                        # convert player id to integer

```

Validate there are exactly 64 * 7 ids (64 players, 7 games per player).

```{r}
length(p_opponent_ids) == 64 * 7
```


We can split opposing player ids into equal sized partitions (7 each). Index into the resulting opponents list is the player id!  

For example, KENNETH  has player id 12 and his opponents will be found at index 12. As expected, we observe that the opponent for game 5 is missing (NA).

```{r}
p_opponents <- split(p_opponent_ids,                   
                     cut(seq_along(p_opponent_ids),
                     length(mPlayersAndGames),
                     labels = FALSE))

ken_pid <- 12              # KENNETH's player id

mPlayersAndGames[ken_pid]  # KENNETH's games

p_opponents[ken_pid]       # KENNETH's opponents
```

## Calculate average rank of opponents for all players

Now we are ready to calculate the average rank of opponents for each player.

    * Sum the ranks of all opponents, ignore NAs
    * Count the number of opponents, ignoring NAs
    * Compute average opponent rank

```{r calculate-avg-opponent-score}
opp_prerating_sum <- map(p_opponents,         # for each player, sum pre-rating of all opponents
                         function(x) sum(player_prerating[x], 
                                         na.rm = TRUE)) 

opp_count <- map(p_opponents, 
                 function(x) sum(!is.na(x)))  # for each player, count opponent ids that are not NA

opp_avg_prerating <- map2(opp_prerating_sum, # compute player average score
                          opp_count, 
                          ~ round(.x / .y, 0)) 
```

Let's validate our transformations. As expected, number of games with opponents, for KENNETH J TACK, VIRAJ MOHILE, and ASHWIN BALAJI are 6, 6, and 1, respectively.  

```{r}
mPlayersAndGames[kva_idx]

d <- unlist(opp_count, use.names = FALSE)         # opponent counts
d[kva_idx]
```

Let's do a detailed validation of games played by KENNETH. Visual inspection shows that for games played by KENNETH, the opponent ids are 42, 33, 5, 38, 1, and 3. Note, the 5^th^ game was a *bye* and has NA opponent id.

```{r}
kenneth_idx <- 12
mPlayersAndGames[kenneth_idx]
```

Ranks of KENNETH opponents are 1332, 1449, 1655, 1423, 1794, and 1384.

```{r}
kenneth_opponents <- c(42, 33, 5, 38, 1, 3)
unlist(mStatesAndRanks[kenneth_opponents])
```

Let's hand compute the sum and average of KENNETH's opponent rank. 

```{r}
kenneth_sum <- sum(1332, 1449, 1655, 1423, 1794, 1384) # sum of opponent ranks
kenneth_sum

kenneth_avg <- round(kenneth_sum / 6, 0)                     # avg opponent rank
kenneth_avg

```

Now, validate the hand computed sum/average against our transformations. Checks out!

```{r}
kenneth_sum == opp_prerating_sum[[kenneth_idx]]
kenneth_avg == opp_avg_prerating[[kenneth_idx]]
```

## Create result data frame

Now let's warp all computed attributes into the result data frame.

```{r create-data-frame}
player_opp_avg_prerating <- unlist(opp_avg_prerating, use.names = FALSE) # unlist

result_df <- data.frame(player_id,        # create result data.frame
                 player_name, 
                 player_state, 
                 player_total_points, 
                 player_prerating, 
                 player_opp_avg_prerating) 

colnames(result_df) <- c('id',              # change column names
                  'name', 
                  'state', 
                  'tot_points', 
                  'pre_rating', 
                  'avg_opponent_pre_rating')
```

Lets take a look at the final results.
```{r}
head(result_df, 5)

NROW(result_df)

result_df[kva_idx, ]
```

As final sanity check, verify that key data frame values for KENNETH match earlier validations.

```{r}
opp_avg_prerating[[kenneth_idx]] == result_df[kenneth_idx, "avg_opponent_pre_rating"]

player_prerating[kenneth_idx] == result_df[kenneth_idx, "pre_rating"]

result_df[kenneth_idx, ]
```

## Generate output CSV

Now we can generate the output CSV file in current working directory.

```{r generate-csv}
write.csv(result_df,              # data
          "player_analysis.csv",  # output file name in current working dir
          row.names = FALSE)      # skip row names since same as player id
```

