---
title: "Project 2 - Solution 2"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Assignment

## Data set from Enid Roman

Original data set at in [salary survey data](https://docs.google.com/spreadsheets/d/1IPS5dBSGtwYVbjsfbaMCYIWnOuRmJcbequohNxCyGVw/edit?resourcekey#gid=1625408792).  

This large data file must be downloaded from  [Github](https://github.com/himalayahall/DATA607/blob/main/Project2/enid_roman.csv) and installed in the current working directory. 

```{r}
df <- read.csv("enid_roman.csv")
str(df)
```

The column headers are descriptive but messy to work with. Let's assign concise column headers.

```{r}
colnames(df) <- c('timestamp', 
                   'age', 
                   'industry', 
                   'title',
                   'title_context',
                   
                   'annual_salary',
                   'additional_comp',
                   'currency',
                   'currency_other',
                   'income_context',
                   
                   'country',
                   'state_usa',
                   'city',
                   'years_experience_overall',
                   'years_experience_in_field',
                   
                   'highest_education',
                   'gender',
                   'race'
                   )
str(df)
```

Next, we observe that **age** is a range. Let's take a look at the unique values in age column. We see that age has two problematic values: **under 18** and **65 or over**.

```{r}
unique(df$age)
```

Let's do the following transformations:

1. Convert **under 18** to 0-18
1. Convert **65 or Over** to 65-150
1. Separate **age** into two columns: age_lower and age_upper

```{r}
df <- df %>%
        mutate(age = ifelse(age == "under 18", "0-18", age)) %>%
        mutate(age = ifelse(age == "65 or over", "65-150", age)) %>%
        separate(age, into = c('age_lower', 'age_upper'), convert = TRUE)
```

Let's take a look at the unique values in **age_lower** and **age_upper** columns. Successful transformation!

```{r}
unique(df$age_lower)
unique(df$age_upper)
```

Now let's take a look at the unique values in columns **years_experience_overall** and **years_experience_in_field**.

```{r}
unique(df$years_experience_overall)
unique(df$years_experience_in_field)
```

There are a number of transformations that we can make to **years_experience_overall** and **years_experience_in_field**:

1. Convert **41 years or more** to **41-100**
1. Convert **1 year or less** to **0-1**
1. Remove whitespace and alphabetic characters from all values
1. Separate **years_experience_overall**: years_experience_overall_lower and years_experience_overall_upper
1. Separate **years_experience_in_field**: years_experience_in_field_lower and years_experience_in_field_upper

```{r}
df <- df %>%
        mutate(years_experience_overall = 
                   ifelse(years_experience_overall == "1 year or less", "0-1", years_experience_overall)) %>%
        mutate(years_experience_overall = 
                   ifelse(years_experience_overall == "41 years or more", "41-100", years_experience_overall)) %>%
        mutate(years_experience_overall = str_remove(years_experience_overall, "years")) %>%
        mutate(years_experience_overall = str_remove_all(years_experience_overall, "\\s+")) %>%
        separate(years_experience_overall, into = c('years_experience_overall_lower', 'years_experience_overall_upper'), convert = TRUE) %>%

        mutate(years_experience_in_field = 
                   ifelse(years_experience_in_field == "1 year or less", "0-1", years_experience_in_field)) %>%
        mutate(years_experience_in_field = 
                   ifelse(years_experience_in_field == "41 years or more", "41-100", years_experience_in_field)) %>%
        mutate(years_experience_in_field = str_remove(years_experience_in_field, "years")) %>%
        mutate(years_experience_in_field = str_remove_all(years_experience_in_field, "\\s+")) %>%
        separate(years_experience_in_field, into = c('years_experience_in_field_lower', 'years_experience_in_field_upper'), convert = TRUE)
```

Verify transformations. Success!

```{r}
unique(df$years_experience_overall_lower)
unique(df$years_experience_overall_upper)
unique(df$years_experience_in_field_lower)
unique(df$years_experience_in_field_upper)
```

A quick look at **country** column shows a number of issues:

1. Uppercase and lowercase names: e.g. 'denmark', 
1. Misspelling - e.g. 'Nederland', 'Englang', 'united stated'
1. Variations - e.g. 'United States of America', 'united states', 'U.s.a'

```{r}
head(df$country, n = 20)
```

Fortunately, we can leverage the **countrycode** library to help clean up the country data.

```{r}
library(countrycode)
df <- df %>%
        mutate(country = countrycode::countryname(country)) %>%
        mutate(country_code = countrycode::countrycode(country, origin = 'country.name', destination = 'iso3c'))
head(df$country, n = 20)

```

Unfortunately, there is a lot of remaining noise in the country data. For example:

1. ARGENTINA BUT MY ORG IS IN THAILAND - there are two countries mentioned
1. Company in Germany. I work rom Pakistan. - again, two countries mentioned

With greater effort, more cleanup of country names is possible. Left as an exercise for another day!

Final cleanup for this data set is to remove number formatting characters from  **annual_salary** and **additional_comp** columns and convert to numeric. Also add **total_comp** column as sum, of annuial_salary and additional_comp.

```{r}
df <- df %>%
    mutate(annual_salary = str_remove_all(annual_salary, "[\\s,]+")) %>%
    mutate(annual_salary = as.numeric(annual_salary)) %>%

    mutate(additional_comp = str_remove_all(additional_comp, "[\\s,]+")) %>%
    mutate(additional_comp = as.numeric(additional_comp)) %>%
    
    mutate(total_comp = annual_salary + additional_comp) 
```

The analysis for this data set is to compare salaries by gender and years of experience. Let's find professions with the largest number of observations so we can make some generalizations.

```{r}
top_n <- df %>% 
            filter(currency == 'USD') %>%
            filter(gender %in% c('Man', 'Woman')) %>%
            filter(years_experience_in_field_lower > 1) %>%
            count(title) %>% 
            arrange(desc(n)) %>%
            head(7) 

top_n
```

To compare salaries by gender and years of experience let's plot a bar chart with the following filters: 

1. USD salaries
1. Man/Woman gender
1. More than 1 year experience in chosen field
1. Selected top-n professions

Plot shows that across the top-n professions average total compensation for men is higher than for women.

```{r}
library(ggpubr)

df %>%
    filter(currency == 'USD') %>%
    filter(gender %in% c('Man', 'Woman')) %>%
    filter(years_experience_in_field_lower > 1) %>%
    filter(title %in% top_n[[1]]) %>%
    group_by(title, gender) %>%
    summarize(avg_total_comp = round(mean(total_comp, na.rm = TRUE), 0)) %>%
    ggbarplot(x = 'title', 
              title = 'Average total comp by gender',
              y = 'avg_total_comp',
              color = 'gender',
              fill = 'gender',
              palette = 'jco',
              label = TRUE, 
              lab.col = "white",
              ab.pos = "in")
```
