---
title: "Project 2 - Solution 4"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magick)
library(tesseract)
```

# Assignment

## Data set

Let's load JPEG image containing the data.

```{r load-data}
img_url <- "file:///Users/jawaidhakim/Desktop/Screen Shot.png"

raw_img <- image_read(img_url)
image_ggplot(raw_img)
```

## Tidy

Let's do a quick OCR using raw image.

1. Column headers are not recognized 
1. Gender for Sally is misread as incorrectly as *a* instead of *1*
1. Age for Michael is read incorrectly as  *The* instead of *"7"*
1. Age for Elizabeth is read incorrectly as *s* instead of *"6"*
1. Gender for Mary is read incorrectly as *4* instead of *1*

```{r}
raw_img %>% 
    ocr()
```

Color images are problematic for OCR engines. Convert image to grayscale.

1. Column headers are correctly recognized 
1. Gender for Sally is read correctly as *1*
1. Age for Michael is read correctly *"7"* but gender is incorrect
1. Age for Elizabeth is read correctly as *"6"s*
1. Gender for Mary is read incorrectly as *4* instead of *1*


```{r}
raw_img %>% 
  image_quantize(colorspace = "Gray") %>% 
  image_threshold(type = "white", threshold = "50%") #%>%
  image_ggplot()
```

OCR grayscale image.

1. Column headers are recognized 
1. Gender for Sally is read correctly as *1*
1. Age for Michael is read incorrectly *wa* instead of *"7"*
1. Age for Elizabeth is correctly as *"6"*
1. Gender for Mary is read incorrectly as *4* instead of *1*

```{r}
raw_img %>% 
  image_quantize(colorspace = "Gray") %>% 
  image_transparent(color = "white", fuzz=20) %>% 
  image_background("white") %>%
  #image_negate() %>%
  ocr()
```


```{r}
fuzz_fun <- function(fuzz){
  raw_img %>% 
    image_quantize(colorspace = "gray") %>% 
    image_transparent(color = "white", fuzz=fuzz) %>% 
    image_contrast(sharpen = 20) %>%
    image_background("white")
}
```

OCR engines are also misled by gray lines. Let's eliminate the image background.

1. Column headers are recognized 
1. Gender for Sally is read correctly as *1*
1. Age for Michael is read incorrectly *wa* instead of *"7"*
1. Age for Elizabeth is correctly as *"6"*
1. Gender for Mary is read correctly as *1*

```{r}
no_grid <- raw_img %>% 
    image_quantize(colorspace = "gray") %>% 
    image_transparent(color = "white", fuzz=30) %>% 
    image_contrast(sharpen = 10) %>%
    image_background("white") 
```

```{r}
no_grid %>%
  image_negate() %>% # negate
  image_morphology(method = "Thinning", kernel = "Rectangle:20x1") %>%
  image_negate() %>% # back to white
  image_ggplot()
```

```{r}
no_grid %>% 
    image_ocr()
```

