---
title: "Project 2 - Solution 3"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Assignment

## Data set from Wilson Ng

Let's load the data set.

```{r}
df <- read.csv("wilson_ng.csv")
df
```

## Tidy

This is a minimal data set with the following untidy characteristics:

1. Contains duplicate entries for fruits - e.g. apple
1. Duplicate items have different prices
1. Currency symbol is embedded in price

Let's cleanup the data as follows:

1. Remove currency symbol from price
1. Convert price to numeric
1. Average price/calories (in case of duplicates)
1. Remove duplicate fruit entries

Let's validate the mutations:

1. Average price of peach is 3.5, which is indeed the average of 3 and 4
1. Average calories for peach is 55 (unchanged), which is also correct

```{r}
df <- df %>%
        group_by(item) %>%
        mutate(price = str_remove(price, '\\$')) %>% # remove $ symbol
        mutate(price = as.numeric(price)) %>%        # convert to numeric
        summarize(price = mean(price),               # average price/calories
                  calories = mean(calories)) %>%
        arrange(item)
head(df)
```

## Analysis

Finally, let's do the analysis and find the cheapest fruit(s) with the most calories. Keeping in mind that there may be more than one qualifying fruit, we first select up to 5 items with the lowest price and then select up to 2 items with the highest calories. 

```{r}
min_df <- df %>%
            slice_min(price, n = 5)     # 5 items

min_df <- min_df %>%                                       
            slice_max(calories, n = 2)  # 2 items
```

We see that **mango** is the cheapest fruit (not in the USA!) with the highest calories. Bannana (misspelt) is a close second with same price but fewer calories.

```{r}
min_df
```


