---
title: "Project 2 - Solution 1"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Assignment

## Data set from Jhalak Das

Let's load CSV file containing the data.

```{r load-data}
df <- read.csv("jhalak_das.csv")
df
```

## Tidy

A quick look at the data frame structure identifies some **wider** data characteristics:

1. Term scores separate columns
1. **sex and age** are concatenated into a single column
1. Personal data of students (name, sex, phone) intermingled with test data

Let's start by separating **sex.and.age** column into **sex** and **age** columns. Note, during separation we also convert **age** into integer. 

```{r separate-cols}
df <- df %>%
    separate('sex.and.age', 
             sep = '_', 
             into = c('sex', 'age'), 
             convert = TRUE)
head(df, n = 20)
```

Next, we **pivot longer** by collapsing the term column names  - term1., term.2, term.3 - into a **term** column and putting values into a **score** column.

```{r pivot-longer}
df <- df %>%
        pivot_longer(cols = 7:9,
                     names_to = "term",     # col with term names 
                     values_to = "score")   # col with scores
head(df)
```

Next, clean up strings: uppercase values, trim white spaces, replace '.' in term column data with space.

```{r trim}
df <- df %>%        
        mutate(name = str_to_upper(str_trim(name))) %>% 
    
        mutate(sex = str_to_upper(str_trim(sex))) %>%     
    
        mutate(test.number = str_to_upper(str_trim(test.number))) %>% 
    
        mutate(term = str_replace(
                        str_to_upper(str_trim(term)), '\\.', ' '))
head(df)
```

Next, we separate the data into two tables. First table contains personal details of students (id, name, phone, sex).

```{r extract-personal-data}
df_personal <- df %>%
            select(id, name, phone, sex) %>%
            distinct(id, .keep_all = TRUE) %>%
            arrange(id)
head(df_personal)
```

Second table contains key details of students and test scores (id, name, test.number, term, score). Note, strictly speaking **id** is sufficient to identify students but  having the **name** in this table is convenient.

That completes the cleanup/analysis of this data set.

```{r extract-test-data}
df_test <- df %>%
            select(id, name, test.number, term, score) %>%
            arrange(id)
head(df_test, n = 10)
```
