---
title: "Assignment9 - Web APIs"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r setup, include=TRUE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(jsonlite)
library(assertive)   # for validating request parameters
```

# New York Time Books API Vignette

In this solution we will use the New York Times [Books API](https://developer.nytimes.com/docs/books-product/1/overview) to retrieve information about The New York Times Best Sellers lists.  

# Create R API

Set the API Key in one place to make it easy to change.

```{r api-access-key}
API_KEY <- "2mGra8XCfEhUdE4WxThgvyCs55BZLfLK"
```

Create utilities to facilitate development of API functions. These include URL encoding (to escape reserved characters), modifying URL placeholders, adding request parameters, and adding API access key.

```{r utility-functions}

#' Finalize URL. Add API access key top URL and URLEncode.
#'
#' @param url URL
#' @param verbose - TRUE to print final URL.
#' @returns URL with API access key and URL  encoding
FinalizeUrl <- function(url, verbose = FALSE) {
  if (str_detect(url, "\\?")) {
    url <- paste0(url, "&")
  } else {
    url <- paste0(url, "?")
  }
  url <- paste0(url, "api-key=", API_KEY)
  url <- URLencode(url)
  if (verbose) {
      print(url)
  }
  return (url)
}

#' URL component substitution and request parameter injection.
#'
#' @param url URL
#' @param url_subs URL substitution dictionary. Keys are placeholders in the URL and values are substitution strings.
#' @param req_params Request parameter dictionary. Keys are parameter names and values are parameter strings.
#' @returns URL with URL substitutions and request parameters.
#' @example AddReqParams("http://foo.com/{book}/index.json", c('{book}:'fiction'),c('param':'123'))
AddReqParams <- function(url, url_subs = NULL, req_params = NULL) {
  if (!is_empty(url_subs)) {
    for (key in names(url_subs)) {
      url <- str_replace(url, key, url_subs[key])
    }
  }

  if (!is_empty(req_params)) {
    first <- TRUE
    for (key in names(req_params)) {
      if (first) {
        url <- paste0(url, "?")
        first <- FALSE
      } else {
        url <- paste0(url, "&")
      }
      url <- paste0(url, key, "=", req_params[key])
    }
  }
  return (url)
}

#' URL component substitution, request parameter injection, API key injection, and URL encoding.
#'
#' @param url URL
#' @param url_subs URL substitution dictionary. Keys are placeholders in the URL and values are substitution strings.
#' @param req_params Request parameter dictionary. Keys are parameter names and values are parameter strings.
#' @returns URL with URL substitutions and request parameters.
#' @example AddReqParamsAndFinalize("http://foo.com/{book}/index.json", c('{book}:'fiction'),c('key':'123'))
AddReqParamsAndFinalize <- function(url, url_subs = NULL, req_params = NULL) {
  url <- AddReqParams(url, url_subs, req_params)
  url <- FinalizeUrl(url)
  return (url)
}
```

API endpoints.

```{r api-endpoints}
BOOKS_API_VERSION <- "v3"

BOOKS_API_BASE_URL <- paste0("https://api.nytimes.com/svc/books", "/", BOOKS_API_VERSION)

BOOKS_API_LIST_NAMES_URL <- paste0(BOOKS_API_BASE_URL, "/lists/names.json")

BOOKS_API_LIST_URL <- paste0(BOOKS_API_BASE_URL, "/lists.json")

BOOKS_API_LIST_BYDATE_URL <- paste0(BOOKS_API_BASE_URL, "/lists/{date}/{list}.json")

BOOKS_API_HISTORY_URL <- paste0(BOOKS_API_BASE_URL, "/lists/best-sellers/history.json")

BOOKS_API_OVERVIEW_URL <- paste0(BOOKS_API_BASE_URL, "/lists/overview.json")

BOOKS_API_FULL_OVERVIEW_URL <- paste0(BOOKS_API_BASE_URL, "/lists/full-overview.json")
```

API functions. Each function encapsulates one New York Times Books API and the required and optional parameters.

```{r api-functions}

#' Get Best Seller List names.
#'
#' https://developer.nytimes.com/docs/books-product/1/routes/lists/names.json/get
API.GetNames <- function(verbose = FALSE) {
  url <- FinalizeUrl(BOOKS_API_LIST_NAMES_URL,verbose)
  o <- fromJSON(url)
  return (o)
}

#' Get Best Seller List.
#'
#' https://developer.nytimes.com/docs/books-product/1/routes/lists.json/get
API.GetNamedList <- function(list_name, bestsellers_date = NULL, published_date = NULL, offset = NULL, verbose = FALSE) {
  url <- BOOKS_API_LIST_URL

  params <- c("list" = list_name)

  if (!is_empty(bestsellers_date)) {
    params["bestsellers-date"] <- bestsellers_date
  }

  if (!is_empty(published_date)) {
    params["published-date"] <- published_date
  }

  if (!is_empty(offset)) {
    assert_all_are_divisible_by(offset, 20) # must be multiple of 20
    params["offset"] <- str(offset)
  }

  url <- AddReqParams(url, NULL, params)
  url <- FinalizeUrl(url, verbose)
  o <- fromJSON(url)
  return (o)
}

#' Get Best Seller list
#'
#' https://developer.nytimes.com/docs/books-product/1/routes/lists/%7Bdate%7D/%7Blist%7D.json/get
API.GetListByDate <- function(list_date = "current", list_name, offset = NULL, verbose = FALSE) {
  url <- BOOKS_API_LIST_BYDATE_URL

  url_subs <- c("\\{date\\}" = list_date, "\\{list\\}" = list_name)

  params <- c()
  if (!is_empty(offset)) {
    assert_all_are_divisible_by(offset, 20) # must be multiple of 20
    params["offset"] <- str(offset)
  }

  url <- AddReqParams(url, url_subs, params)
  url <- FinalizeUrl(url, verbose)
  o <- fromJSON(url)
  return (o)
}

#' Get Best Seller list history.
#'
#' https://developer.nytimes.com/docs/books-product/1/routes/lists/best-sellers/history.json/get
API.GetHistory <- function(age_group = NULL, 
                           author = NULL, 
                           contributor = NULL, 
                           isbn = NULL, 
                           offset = NULL, 
                           price = NULL, 
                           publisher = NULL, 
                           title = NULL, 
                           verbose = FALSE) {
  url <- BOOKS_API_HISTORY_URL

  params <- c()

  if (!is_empty(age_group)) {
    params["age-group"] <- age_group
  }

  if (!is_empty(author)) {
    params["author"] <- author
  }

  if (!is_empty(contributor)) {
    params["contributor"] <- contributor
  }

  if (!is_empty(isbn)) {
    params["isbn"] <- isbn
  }

  if (!is_empty(offset)) {
    assert_all_are_divisible_by(offset, 20) # must be multiple of 20
    params["offset"] <- offset
  }

  if (!is_empty(price)) {
    params["price"] <- price
  }

  if (!is_empty(age_group)) {
    params["age-group"] <- age_group
  }

  if (!is_empty(publisher)) {
    params["publisher"] <- publisher
  }

  if (!is_empty(title)) {
    params["title"] <- title
  }

  url <- AddReqParams(url, NULL, params)
  url <- FinalizeUrl(url, verbose)
  o <- fromJSON(url)
  return (o)
}

#' Get all books for all the Best Sellers lists for specified date.
#'
#' https://developer.nytimes.com/docs/books-product/1/routes/lists/full-overview.json/get
API.GetFullOverview <- function(published_date = NULL, verbose = FALSE) {
  url <- BOOKS_API_FULL_OVERVIEW_URL

  params <- c()
  if (!is_empty(published_date)) {
    params["published_date"] <- published_date
  }

  url <- AddReqParams(url, NULL, params)
  url <- FinalizeUrl(url, verbose)
  o <- fromJSON(url)
  return (o)
}

#' Get all books for all the Best Sellers lists for specified date.
#'
#' https://developer.nytimes.com/docs/books-product/1/routes/lists/overview.json/get
API.GetOverview <- function(published_date = NULL, verbose = FALSE) {
  url <- BOOKS_API_OVERVIEW_URL

  params <- c()
  if (!is_empty(published_date)) {
    params["published_date"] <- published_date
  }

  url <- AddReqParams(url, NULL, params)
  url <- FinalizeUrl(url, verbose)
  o <- fromJSON(url)
  return (o)
}

```

# API Test

Time to put the APIs through their paces!  

## Names of all Best Seller lists

```{r best-seller-list-names}
ret <- API.GetNames(verbose = TRUE)
if (ret$status == "OK") {
    #str(ret)
    head(ret$results, n = 2)
}
```

## Best Seller list 

Get list `Advice How-To and Miscellaneous`. List name has features (whitespace, `-`) that test URL encoding.

```{r best-seller-list-1}
ret <- API.GetNamedList(list_name = 'Advice How-To and Miscellaneous', verbose = TRUE)
if (ret$status == "OK") {
    #str(ret)
    head(ret$results, n = 2)
}
```

## Best Seller list by date

```{r best-seller-list-2}
ret <- API.GetListByDate(list_name = 'education', verbose = TRUE)
if (ret$status == "OK") {
    #str(ret)
    head(ret$results$books, n = 2)
}
```

## Best Seller list history

### By age group

```{r best-seller-list-history-1}
ret <- API.GetHistory(age_group = 15, verbose = TRUE)
if (ret$status == "OK") {
    #str(ret)
    head(ret$results, n = 2)
}
```

### By age group and publisher

```{r best-seller-list-history-2}
ret <- API.GetHistory(age_group = 15, publisher = 'Bloomsbury', verbose = TRUE)
if (ret$status == "OK") {
    #str(ret)
    head(ret$results)
}
```

### By publisher and price

```{r best-seller-list-history-3}
ret <- API.GetHistory(publisher = 'HarperCollins', verbose = TRUE)
if (ret$status == "OK") {
    #str(ret)
    head(ret$results, n = 2)
}
```

### By publisher and offset (offset must be a multiple of 20)

```{r best-seller-list-history-4}
ret <- API.GetHistory(publisher = 'HarperCollins', offset = 20, verbose = TRUE)
if (ret$status == "OK") {
    #str(ret)
    head(ret$results, n = 2)
}
```

## All books for all the Best Sellers lists for specified date

```{r best-seller-list-overview}
ret <- API.GetOverview(published_date = '2013-05-15', verbose = TRUE)
if (ret$status == "OK") {
    print(ret$num_results)
}
```

