---
title: "Assignment 06 - Working with Tidy Data"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Assignment

The chart above describes arrival delays for two airlines across five destinations. Your task is to:
1. Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above.
You’re encouraged to use a “wide” structure similar to how the information appears above, so
that you can practice tidying and transformations as described below.
1. Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy
and transform your data.
1.  Perform analysis to compare the arrival delays for the two airlines.
1. Your code should be in an R Markdown file, posted to rpubs.com, and should include narrative
descriptions of your data cleanup work, analysis, and conclusions.

# Solution

## Data Preparation

Let's load a CSV file containing the data and format described above.

```{r load-data}
df <- read.csv("https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%2006%20-%20Working%20eith%20Tidy%20Data/airline_delays.csv")
```

A quick look at the data frame structure shows that the CSV contains two unnamed columns, which have been labeled X and X.1 in the data frame.

```{r}
str(df)
```

Let's rename these columns as *airline* and *arrival_status*, respectively.

```{r rename-cols}
df <- df %>% rename('airline' = 1, 'arrival_status' = 2)
str(df)
```

Looking at the data, we make two observations: (a) there is an (almost) empty row between the two airlines and (b) airline name is missing on **delayed** rows.

```{r ds-head}
head(df)
```

Based on above data observations, first let's remove the empty.

```{r remove-empty-row-and-fill-missing}
df <- df %>%
            filter(! is.na(arrival_status) &  # remove rows with NA or empty status
                     str_length(arrival_status) > 0) 
head(df)
```

Lastly, let's fill in the airline name on **delayed** rows: replace empty **airline** values with NA and fill in missing values **down** the airline column.

```{r}
df <- df %>%
        mutate(airline = ifelse(airline == "", NA, airline)) %>%
        fill(airline, .direction = "down") # fill in missing values
head(df)
```

Next, we observe that data frame in a **wide** format. Specifically, destination cities are given as columns. Let's convert this to a **longer** format.

```{r pivot-longer}
df <- df %>%
            pivot_longer(!c("airline", "arrival_status"), # pivot all cols EXCEPT airline and arrival_status
                        names_to = "dest",                # destination names 
                        values_to = "delay_count")              # destination late flight values
head(df)
```


## Analysis to compare the arrival delays for the two airlines

Let's extract delayed flight data.

```{r extract-delayed-flight-data}
delayed_flights <- df %>%
                    filter(df$arrival_status == "delayed")
delayed_flights
```

Let's delayed flight data as a histogram. Visual inspection shows that AM WEST has greater number of delayed flights to destinations compared to Alaska. The one exception destination is Seattle where Alaska has more delayed flights.

```{r barchart-delayed}
    ggp <- ggplot(data=delayed_flights, aes(x=dest, y=delay_count, fill=airline))
    ggp <- ggp +  ggtitle('Delayed Flights') + theme(plot.title = element_text(hjust = 0.5))
    ggp <- ggp + geom_text(aes(label=delay_count), vjust=-0.2,
                            position = position_dodge(0.9), size=3.5) +
                            scale_fill_brewer(palette="Paired") +
            geom_bar(stat="identity", position=position_dodge()) 
ggp
```


