---
title: "R Character Manipulation and Date Processing"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# Assignment

## Search for majors containing "DATA" or "STATISTICS" - 1

Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"  

Load data from GitHub *raw* file.

```{r load-from-github}
ds <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/2d2ff3e9457549d51f8e571c52099bfe9b2017ad/college-majors/majors-list.csv")
```

Let's take a look at the unique values of *Major* and *Major_Category* columns. There are 174 unique majors and 16 major categories.

```{r}
glimpse(as.factor(ds$Major))
glimpse(as.factor(ds$Major_Category))
```

Let's assume the search terms may appear in either Major or Major Category columns and we want to consider both. 

Filter data using *grepl* regular expressions, one per column. The search terms are found in 3 majors: 

* COMPUTER PROGRAMMING AND DATA PROCESSING
* MANAGEMENT INFORMATION SYSTEMS AND STATISTICS
* STATISTICS AND DECISION SCIENCE

```{r}
ds %>%
    filter(
            grepl("DATA|STATISTICS", ignore.case = TRUE, Major) | 
            grepl("DATA|STATISTICS", ignore.case = TRUE, Major_Category)) %>%
    arrange(Major)
```

## Transform data - 2

Write code that transforms the data below:
[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry"
Into a format like this:
c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")  

 First we generate the input data as a string.

```{r}
input_data <- '[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"
[9] "elderberry"   "lime"         "lychee"       "mulberry" 
[13] "olive"        "salal berry"'
input_data
```

Extract fruit names from the input data using *str_extract_all* and flatten the result into a vector using *unlist*.

```{r}
input_split <- unlist(str_extract_all(input_data, '"[a-zA-Z ]+"'))
input_split
```

All that remains to be done is removing those pesky double quotes using *str_remove*. And we are finished!

```{r}
str_remove_all(input_split, '"')
```

Of course, we can *pipe* the two transformations for brevity and still get same results.

```{r}
unlist(str_extract_all(input_data, '"[a-zA-Z ]+"')) %>% str_remove_all('"')
```

## Describe Data - 3

1. (.)\1\1 : matches strings of length 3 where all characters are identical. Example, aaa, BBB, 111, etc.
1. (.)(.)\2\1 : *palindromes* of length 4. For example, abba, 1221, etc.
1. (..)\1 : matches strings of length 4 where the first 2 characters are repeated. For example, abab, 1212.
1. (.).\1.\1 : matches strings of length 5 where the 1^st^, 3^rd^, 5^th^ characters are the same. Example, abaka, 12141, etc.
1. (.)(.)(.).\*\3\2\1 : matches strings of length 6 or greater where the first 3 characters are repeated at the end of the string in reverse order. For example, abccba, abc123cba.

## Construct regular expressions to match words - 4

1. Start and end with same character: (.).\*\1
1. Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.): .\*([a-zA-Z])([a-zA-Z]).\*\1\2.\*
1. Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.): .\*([a-zA-Z]).\*\1.\*\1.\*
