---
title: "Untitled"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


The chart above describes arrival delays for two airlines across five destinations. Your task is to:
(1) Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above.
You’re encouraged to use a “wide” structure similar to how the information appears above, so
that you can practice tidying and transformations as described below.
(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy
and transform your data.
(3) Perform analysis to compare the arrival delays for the two airlines.
(4) Your code should be in an R Markdown file, posted to rpubs.com, and should include narrative
descriptions of your data cleanup work, analysis, and conclusions. Please include in your
homework submission:

```{r load-data}
input_ds <- read_csv("airline_delays.csv")
str(input_ds)
```
```{r rename-cols}
input_ds <- input_ds %>% rename(airline = 1, arrival_status = 2)
```


```{r view-ds-structure}
str(input_ds)
```

```{r ds-head}
head(input_ds)
```
```{r remove-empty-row-and-fill-missing}
input_ds <- input_ds %>% 
            filter(! is.na(arrival_status)) %>%
            fill(airline)
head(input_ds)
```

```{r pivot-longer}
input_ds <- input_ds %>%
                pivot_longer(!c("airline", "arrival_status"), 
                            names_to = "dest", 
                            values_to = "count")
head(input_ds)
```


```{r extract-delayed-flight-data}
delayed_flights <- input_ds %>%
                    filter(input_ds$arrival_status == "delayed")
delayed_flights
```

```{r barchart-delayed}
    ggp <- ggplot(data=delayed_flights, aes(x=dest, y=count, fill=airline))
    ggp <- ggp +  ggtitle('Delayed Flights') + theme(plot.title = element_text(hjust = 0.5))
    ggp <- ggp + geom_text(aes(label=count), vjust=-0.2,
                            position = position_dodge(0.9), size=3.5) +
                            scale_fill_brewer(palette="Paired") +
            geom_bar(stat="identity", position=position_dodge()) 
ggp
```


