---
title: "Assignment - Working with XML and JSON in R"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(jsonlite)
library(xml2)
library(XML)
library(htmltab)
```

# Assignment
    
>Pick three of your favorite books on one of your favorite subjects. At least one of the books should have more
than one author. For each book, include the title, authors, and two or three other attributes that you find
interesting.    
Take the information that you’ve selected about these three books, and separately create three files which
store the book’s information in HTML (using an html table), XML, and JSON formats (e.g. “books.html”,
“books.xml”, and “books.json”). To help you better understand the different file structures, I’d prefer that you
create each of these files “by hand” unless you’re already very comfortable with the file formats.
Write R code, using your packages of choice, to load the information from each of the three sources into
separate R data frames. Are the three data frames identical?  
Your deliverable is the three source files and the R code. If you can, package your assignment solution up into
an .Rmd file and publish to rpubs.com. [This will also require finding a way to make your three text files
accessible from the web]

# Data

Each book entry in the data set contains 3 books, each with 5 variables: title, author, medium, pages, and isbn. At least one book has 2 authors. For example:

>       {
        "title": "All the Ways We Said Goodbye",
        "author": "Lauran Willig",
        "author": "Karen White",
        "medium": "Hardcover",
        "pages": 448,
        "isbn": "0062931091"
      } 
        
## JSON

Let's use **jsonlite** to load the same data. As expected, the resulting data frame contains 5 observations. Furthermore, we observe *pages* variable is of *int* type. A nice verification that JSON data types have been correctly imported. Multiple authors were modeled as a String Array in JSON and have been read in as comma-separated string. 4 observation of 5 variables. Not bad!

```{r}
d <- jsonlite::fromJSON(txt = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.json", simplifyDataFrame = TRUE)
df_json2 <- as.data.frame(d)
str(df_json2)
df_json2
```

## XML

Next, we load the same data in XML format and run into an immediate issue with multiple authors; the call to *xmlToDataFrame* fails with an error indicating that multiple elements with the same name are not allowed. Note, the offending [XML](https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books2.xml) contains 2 *author* elements.  

To get around this issue we have two choices - (a) add second author using a new tag, e.g.*author2* or (b) add both authors as a single comma-separated string. Since a comma-separated string is similar to how above JSON was read in, that's what we will do. Note, the *pages* variable is of type *char*. 4 observation of 5 variables. 

```{r}
book_data_xml <- xml2::read_xml("https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.xml")
book_xml <- XML::xmlParse(book_data_xml)
df_xml <- xmlToDataFrame(nodes=getNodeSet(book_xml, "//book"))
str(df_xml)
df_xml
```

To address the issue of type safety in XML documents one can leverage **XML Schema** (DTD or XSD). XML Schema is a very powerful mechanism for defining the structure of XML documents, type safety, namespace, etc. For example, here is a basic [XSD](https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.xsd) for this assignment. Note, this schema describes the various XML elements, associated data types, cardinality, namespace, etc. This schema is just scratching the surface of what is possible. The XSD can be used by validating XML parsers to check whether XML documents are well-formed and also to correctly parse elements and data types.  

Let's validate the XML document against the Schema XSD. Validation succeeds! 

```{r validate-xml}
doc <- read_xml("https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.xml", package = "xml2")

xsd <- read_xml("https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.xsd", package = "xml2")

xml_validate(doc, xsd)
```


## HTML

Finally, the same data is loaded in HTML format. Just like the XML above, we encounter an issue with duplicate authors. Although this time the parsing completes, the data is corrupted - name of second author is put in the *medium* variable and variables further to the right are corrupted.  

```{r}
df_html <- htmltab::htmltab(doc = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.html")
str(df_html)
df_html
```

To get around this issue, we follow the same strategy as with XML above and put author names as comma-separated values. 4 observation of 5 variables. It works!

```{r}
df_html <- htmltab::htmltab(doc = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books2.html")
str(df_html)
df_html
```


# Conclusion

We have loaded JSON, XML, and HTML data. All three are *self-describing* (human readable), hierarchical (nested structures), and supported by most programming languages. JSON is perhaps the most widely used format among the three and provides out of the box support for basic data types (string, integer, numeric, bool). And, although XML/XSD is more powerful and has greart support for data modeling and type safety, it needs much more time and effort to bootstrap.

