---
title: "Assignment 8 - Working with XML and JSON in R"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(jsonlite)
library(xml2)
library(XML)
library(htmltab)
```

# Assignment
    
>Pick three of your favorite books on one of your favorite subjects. At least one of the books should have more
than one author. For each book, include the title, authors, and two or three other attributes that you find
interesting.  
Take the information that you’ve selected about these three books, and separately create three files which
store the book’s information in HTML (using an html table), XML, and JSON formats (e.g. “books.html”,
“books.xml”, and “books.json”). To help you better understand the different file structures, I’d prefer that you
create each of these files “by hand” unless you’re already very comfortable with the file formats.
Write R code, using your packages of choice, to load the information from each of the three sources into
separate R data frames. Are the three data frames identical?  
Your deliverable is the three source files and the R code. If you can, package your assignment solution up into
an .Rmd file and publish to rpubs.com. [This will also require finding a way to make your three text files
accessible from the web]

# Data

Each book entry in the data set contains 3 books, each with 5 variables: title, author, medium, pages, and isbn. At least one book has 2 authors. For example:

>       {
        "title": "All the Ways We Said Goodbye",
        "author": "Lauran Willig",
        "author": "Karen White",
        "medium": "Hardcover",
        "pages": 448,
        "isbn": "0062931091"
      } 
        
# JSON

Let's use **jsonlite** to load the same data. As expected, the resulting data frame contains 5 observations. Furthermore, we observe *pages* variable is of *int* type. A nice verification that JSON data types have been correctly imported.

```{r}
d <- jsonlite::fromJSON(txt = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.json", simplifyDataFrame = TRUE)
df_json2 <- as.data.frame(d)
str(df_json2)
df_json2
```

# XML

Next, we load the same data in XML format and run into an immediate issue with multiple authors where the call to *xmlToDataFrame* fails. The XML file used in this step in [here] {https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books2.xml} and contains 2 *author* elements.  

To get around this issue we do the following - add second author using *author2* tag. As expected, the resulting data frame has an additional *author2* column. And the *pages* variable is of type *char*.

```{r}
book_data_xml <- xml2::read_xml("https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.xml")
book_xml <- XML::xmlParse(book_data_xml)
df_xml <- xmlToDataFrame(nodes=getNodeSet(book_xml, "//book"))
str(df_xml)
df_xml
```

To correct the issue of type safety in XML documents one can leverage **XML Schema** (DTD or XSD). XML Schema is a very powerful mechanism for defining the structure of XML documents, type safety, namespace, etc  

For example, here is a basic [XSD](https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.xsd) for this assignment. Note, this schema describes the various XML elements, associated data types, cardinality, etc. This is just scrating the surface of what is possible through XSDs. The XSD can be used by validating XML parsers to check whether XML documents are well-formed and also to correctly parse elements and data types. Unfortunately, I have not (yet) been able to find a XML reader in R that leverages XSD to import correct data types.

# HTML

Finally, the same data is loaded in HTML format. Just like XML, we encounter an issue with duplicate authors. In this case the parsing succeeds but the data is corrupted - get name of second author is put in variable *medium* and the following variables are also corrupted.  


```{r}
df_html <- htmltab::htmltab(doc = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.html")
str(df_html)
df_html
```

To get around this issue we put names of both authors in same variable as comma-separated values. This is less than ideal but it works!

```{r}
df_html <- htmltab::htmltab(doc = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books2.html")
str(df_html)
df_html
```


# Conclusion

We have loaded JSON, XML, and HTML data. All three are *self-describing* (human readable), hierarchical (nested structures), and supported by most programming languages. JSON is perhaps the most widely used format among the three and provides out of the box support for basic data types (string, integer, numeric, bool). And, although XMl/XSD is more powerful and has  greater support for type safety , it needs a lot more time and effort to bootstrap.

