---
title: "Assignment 8 - Working with XML and JSON in R"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
  
  pdf_document: 
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
boxlinks: true
urlcolor: blue
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(jsonlite)
library(rjson)
library(xml2)
library(XML)
library(htmltab)
```

# Assignment
    
>Pick three of your favorite books on one of your favorite subjects. At least one of the books should have more
than one author. For each book, include the title, authors, and two or three other attributes that you find
interesting.  
Take the information that you’ve selected about these three books, and separately create three files which
store the book’s information in HTML (using an html table), XML, and JSON formats (e.g. “books.html”,
“books.xml”, and “books.json”). To help you better understand the different file structures, I’d prefer that you
create each of these files “by hand” unless you’re already very comfortable with the file formats.
Write R code, using your packages of choice, to load the information from each of the three sources into
separate R data frames. Are the three data frames identical?  
Your deliverable is the three source files and the R code. If you can, package your assignment solution up into
an .Rmd file and publish to rpubs.com. [This will also require finding a way to make your three text files
accessible from the web]

# Data

Each book entry in the data set contains 3 books, each with 5 variables: title, author, medium, pages, and isbn. For example:

> "title": "Thinking Fast and Slow",
  "author": "Daniel Kahneman",
  "medium": "Paperback",
  "pages": 499,
  "isbn": "0374533555"
        
# JSON

For our first attempt, we will use **rjson** to load the JSON file. Looking at the structure of the resulting data frame we see that it has 1 observation of 15 variables. This is certainly a surprise as we were expecting 3 observations with 5 variables each.

```{r}
d <- rjson::fromJSON(file = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.json")
df_json <- as.data.frame(d)
str(df_json)
df_json
```

Let's use **jsonlite** to load the same data. Now, as expected, the resulting data frame contains 3 observations of 5 variables. Furthermore, we observe *pages* variable is of *int* type. A nice verification that JSON data types have been correctly imported.

```{r}
d <- jsonlite::fromJSON(txt = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.json")
df_json <- as.data.frame(d)
str(df_json)
df_json
```

# XML

Next, we load the same data in XML format. Looking at the structure of the new data frame, we see that it similar to the JSON data frame. However, unlike the JSON version, the data type for *pages* is *chr* (string). In other words, the *JSON data type has been ignored by the XML reader.

```{r}
book_data_xml <- xml2::read_xml("https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.xml")
book_xml <- XML::xmlParse(book_data_xml)
df_xml <- xmlToDataFrame(nodes=getNodeSet(book_xml, "//book"))
str(df_xml)
df_xml
```

To correct the issue of type safety in XML documents one can leverage **XML Schema** (DTD or XSD). XML Schema is a very powerful mechanism for defining the structure of XML documents, type safety, namespace, etc  

For example, here is a basic [XSD](https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.xsd) for this assignment. Note, this schema describes the various XML elements, associated data types, cardinality, etc. This is just scrating the surface of what is possible through XSDs. The XSD can be used by validating XML parsers to check whether XML documents are well-formed and also to correctly parse elements and data types. Unfortunately, I have not (yet) been able to find a XML reader in R that leverages XSD to import correct data types.

# HTML

Finally, the same data is loaded in HTML format. Here, we see that the data frame is identical to XML and the data type for *pages* is *chr* (string).

```{r}
df_html <- htmltab::htmltab(doc = "https://raw.githubusercontent.com/himalayahall/DATA607/main/Assignment%208/books.html")
str(df_html)
df_html
```

# Conclusion

We have loaded JSON, XML, and HTML data. All three are *self-describing* (human readable), hierarchical (nested structures), and supported by most programming languages. JSON is perhaps the most widely used format among the three and provides out of the box support for basic data types (string, integer, numeric, boolean). Although XMl/XSD is more powerful and has  greater support for type safety , it needs a lot more time and effort to bootstrap.

