---
title: "Untitled"
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

Looks at NBA [players page](https://hoopshype.com/salaries/players/). Hover with mouse over salary table, right-click mouse, select **Copy/Copy CSS**. This is the CSS to extract the salary table.

```{r}
css_selector <- ".hh-salaries-ranking-table"

```

Navigate to the page and grab the salary table using the CSS selector.

```{r}
  salary_table <- read_html("https://hoopshype.com/salaries/players/") %>% 
                    html_element(css = css_selector) %>%
                     html_table()
  salary_table
```
Get first row.

```{r}
first_row <- salary_table[1, ]
first_row
```
Use extracted first row to set column names.

```{r}
colnames(salary_table) <- first_row[1,]
salary_table
```
Remove NA column and the first row (which is now the header).

```{r}
salary_table <- salary_table %>% select(-'NA') %>% 
                                 slice(-1)
salary_table
```
Grab column names.

```{r}
all_cols <- colnames(salary_table)
all_cols
```


Extract yearly salary column names.

```{r}
salary_year_cols <- all_cols[2:length(year_cols)]
salary_year_cols
```

Pivot longer so each salary is in its own row.

```{r}
salary_table <- salary_table %>% 
                    pivot_longer(cols = all_of(salary_year_cols), 
                                 names_to = 'year', 
                                 values_to = 'salary')
salary_table
```
Clean salary (remove comma and currency symbol) and convert to numeric.

```{r}
salary_table <- salary_table %>% 
                 mutate(salary = as.character(gsub(",", "", salary))) %>%
                 mutate(salary = as.numeric(gsub("\\$", "", salary)))
salary_table
```

